\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity}

### 2013-12-08 docs/paper/paper.tex
stukje over EMS aangevuld, pseudocode af

\def \orightarrow {\circ\hspace{-0.42em}\rightarrow}
\def \oleftarrow {\leftarrow\hspace{-0.42em}\circ}
\def \orightline {\circ\hspace{-0.16em}-}
\def \oleftline {-\hspace{-0.16em}\circ}
\def \oline {\circ\hspace{-0.38em}-\hspace{-0.4em}\circ}
\def \srightarrow {\text{\textasteriskcentered}\hspace{-0.62em}\rightarrow}
\def \sleftarrow {\leftarrow\hspace{-0.62em}\text{\textasteriskcentered}}
\def \srightline {\text{\textasteriskcentered}\hspace{-0.16em}-}
\def \sleftline {-\hspace{-0.16em}\text{\textasteriskcentered}}
\def \sline {\text{\textasteriskcentered}\hspace{-0.38em}-\hspace{-0.4em}\text{\textasteriskcentered}}

@@

\subsection{Informal description}
\cite{?}
For finding directed edges, the PC-algorithm first finds the undirected structure.
For each two nodes $X$ and $Y$ which are not neighbours in this structure, it also finds a minimal separating set, which is the smallest set acting as a separated set.
This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.

@@

The first part of the algorithm finds an undirected graph $G$. For each pair of points which are not connected in this graph, it also finds a separating set:

\begin{algorithmic}
\Function{Structure\_PC}{$data$ used to test independence}
\State connect all pairs of point in undirected graph $G$.
\State $n \gets 0$
\While{there are connected pairs $X$, $Y$ such that $n < |\operatorname{Adjecencies}(X)\setminus\{Y\}| $}
  \ForAll{such connected pairs $X$, $Y$}
    \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
      \State test whether $X$ and $Y$ are d-separated given $S$ (Fisher-Z-test on $data$)
      \If{$X$ and $Y$ are d-separated given $S$}
        \State remove the edge $X - Y$ from G
        \State record $S$ as the separating set of $X$ and $Y$
        \State continue with the next pair $X$, $Y$
      \EndIf
    \EndFor
  \EndFor
  n \gets n+1
\EndWhile
\Return G and the separating sets
\EndFunction
\end{algorithmic}

The second part orients edges in this graph $G$ based on the separating sets:

\begin{algorithmic}
\Function{Directional\_PC}{$G$, $sepsets$}
 \State $\text{PDAG} \gets G$
  \ForAll{triples $X$, $Y$, $Z$ with the structure $X - Y - Z$ in $G$ (where $X$ and $Z$ are not connected)}
    \If{$Y$ is not in the separating set of $X$ and $Z$}
      \State Orient $X - Y - Z$ as $X \rightarrow Y \leftarrow Z$ in $\text{PDAG}$
    \EndIf
  \EndFor
  \Repeat
    \ForAll{ordered, connected pairs $X$, $Y$}
      %\If{there is a node $Z$ such that }

      %\EndIf
    \EndFor
  \Until{no more edges are oriented}
\EndFunction
\end{algorithmic}

\ENDWHILE
%\end{algorithm}

>>>>

\def \orightarrow {\circ\hspace{-0.42em}\rightarrow}
\def \oleftarrow {\leftarrow\hspace{-0.42em}\circ}
\def \orightline {\circ\hspace{-0.16em}-}
\def \oleftline {-\hspace{-0.16em}\circ}
\def \oline {\circ\hspace{-0.38em}-\hspace{-0.4em}\circ}
\def \srightarrow {\text{\textasteriskcentered}\hspace{-0.62em}\rightarrow}
\def \sleftarrow {\leftarrow\hspace{-0.62em}\text{\textasteriskcentered}}
\def \srightline {\text{\textasteriskcentered}\hspace{-0.16em}-}
\def \sleftline {-\hspace{-0.40em}\text{\textasteriskcentered}}
\def \sline {\text{\textasteriskcentered}\hspace{-0.38em}-\hspace{-0.4em}\text{\textasteriskcentered}}
\def \soline {\text{\textasteriskcentered}\hspace{-0.16em}-\hspace{-0.4em}\circ}
\def \osline {\circ\hspace{-0.38em}-\hspace{-0.16em}\text{\textasteriskcentered}}

@@

\subsection{Informal description}
The PC algorithm as used for these experiments is the version which is modified to handle latent variables, as also described by Spirtes, Glymour and Scheines \cite{spirtes2000}, pages 165-167.
For finding directed edges, the PC-algorithm first finds the undirected structure.
For each two nodes $X$ and $Y$ which are not neighbours in this structure, it also finds a minimal separating set, which is the smallest set acting as a separated set.
This is done by first assuming all nodes are connected, and then testing all pairs of nodes for independence.

@@

The first part of the algorithm finds an undirected graph $G$. For each pair of points which are not connected in this graph, it also finds a separating set. $\operatorname{Adjacencies} (X)$ is defined as the set of all neighbours of $X$ in the graph G. This graph changes during the algorithm and so does $\operatorname{Adjacencies} (X)$.

\begin{algorithmic}
\Function{Structure\_PC}{$data$}
\State connect all pairs of point in undirected graph $G$.
\State $n \gets 0$
\While{there are connected pairs $X$, $Y$ such that $n < |\operatorname{Adjecencies}(X)\setminus\{Y\}| $}
  \ForAll{such connected pairs $X$, $Y$}
    \ForAll{subsets $S$ of $\operatorname{Adjacencies(X)}$ for which $|S| = n$}
      \State test whether $X$ and $Y$ are d-separated given $S$ (Fisher-Z-test on $data$)
      \If{$X$ and $Y$ are d-separated given $S$}
        \State remove the edge $X - Y$ from G
        \State record $S$ as the separating set of $X$ and $Y$
        \State continue with the next pair $X$, $Y$
      \EndIf
    \EndFor
  \EndFor
  \State $n \gets n+1$
\EndWhile
\Return $G$ and the separating sets
\EndFunction
\end{algorithmic}

The second part orients edges in this graph $G$ based on the separating sets. In this pseudocode, an edge with a star at the end means that that end may be either an arrowhead, an "o" or an empty mark. When an edge is oriented as an edge with a star at the end, this denotes that this end is not changed.

\begin{algorithmic}
\Function{Directional\_PC}{$G$, $sepsets$}
  \State $\text{PDAG} \gets G\text{, with all connections replaced by }\oline$
  \Comment{Find and orient V-structures}
  \ForAll{triples $X$, $Y$, $Z$ with the structure $X\sline Y \sline Z$ in $G$ (where $X$ and $Z$ are not connected)}
    \If{$Y$ is not in the separating set of $X$ and $Z$}
      \State Orient $X \sline Y \sline Z$ as $X \srightarrow Y \sleftarrow Z$ in $\text{PDAG}$
    \EndIf
  \EndFor
  \Repeat
    \ForAll{ordered, connected pairs $X$, $Y$}
      \Comment{Rule 1}
      \If{$X \osline Y$ or $X \sleftline Y$, and there is a node $Z$ such that $Z \srightarrow X$}
        \State Orient $X \sline Y$ as $X \rightarrow Y$
      \EndIf
      \Comment{Rule 2}
      \If{there is a directed path from $X$ to $Y$}
        \State Orient $X \sline Y$ as $X \srightarrow Y$
      \EndIf
    \EndFor
  \Until{no more edges are oriented}
  \State \Return PDAG
\EndFunction
\end{algorithmic}
